---
title: "Effects of Environmental Events on the US population and economy"
author: "Louis Tham"
date: "Tuesday, June 16, 2015"
output:
  html_document:
    keep_md: yes
    self_contained: no
  pdf_document: default
---
####"Reproduceable Research Assignment 2"

### Background
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

### Objective of this Report
The objective of this report is to answer the questions below:

1) Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2) Across the United States, which types of events have the greatest economic consequences?

### The Data
The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:
[Storm Data [47Mb]](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

[National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

[National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.


### 1) Preparation of the Data for Analysis

####a. Prepare the environment
Clear the environment and bring in all relevant libraries
```{r Prepare}
## Remove everything within the environment
rm(list = ls())

library(plyr)
library(dplyr)
library(ggplot2)
#library(lubridate)
library(tidyr)


```


####b. Extraction of Data
1) Prepare the folder, if not available, create the folder
2) Check for the relevant file, if not available, download the file
3) Read in the file into the environment
```{r Extract}

## Cheeck if the folder "Explore_data" exist in the working directory, if not available, create it
if (!file.exists("Reproduce_Data")) {
        dir.create("Reproduce_Data")
}

## Cheeck if the file  "repdata-data-StormData.csv.bz2" exist in the defined folder, if not available, download it
if (!file.exists("Reproduce_Data/repdata-data-StormData.csv.bz2")) {
        fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
        download.file(fileUrl, destfile = "./Reproduce_Data/repdata-data-StormData.csv.bz2")
}

## Read files
data <- read.csv("./Reproduce_Data/repdata-data-StormData.csv.bz2")

```


####c. Basic Transformation of the data
1) To remove typical variation in the primary key
        - Set the data in the primary key to upper case
        - Remove any unnecessary double or triple spaces 
        - Remove any punctuation
2) Group the data based on the primary key (EVTYPE)
        - Only relevant fields such as fatalities, injuries and property damages are extracted
        - All irrelvant data are discarded
3) Remove data that are not significant to this study
        - This study is to identify 2 key points:
                - Events that are most harmful to population health (ie deaths/fatalities) 
                - Events that has the highest economic impact (ie property damages) 
        - Events that results in no fatalities, injuries or property damage are removed

```{r Transform Basic}
# Group data by event and summarize for variables of interest

data$EVTYPE = toupper(data$EVTYPE)
data$EVTYPE = gsub("  ", " ",data$EVTYPE)
data$EVTYPE = gsub("[[:punct:]]", " ", data$EVTYPE)

trim <- function (x) gsub("^\\s+|\\s+$", "", x)

data$EVTYPE = trim(data$EVTYPE)

data = group_by(data, EVTYPE)
data_event = summarise(data, 
                       FATALITIES = sum(FATALITIES), 
                       INJURIES = sum(INJURIES), 
                       DAMAGES = sum(PROPDMG))

## Remove old dataset to free up memory
rm(data)

## Remove irrelevant data
data_event$cumm = data_event$FATALITIES + data_event$INJURIES + data_event$DAMAGES
data_event_tidy = filter(data_event, cumm != 0)
data_event_tidy = data_event_tidy[,1:4]

## Remove old dataset to free up memory
rm(data_event)

```


####d. Reclassification of the data (Cleaning the data)
1) Define the key events as identified by the National Weather Service, documented in the Storm Data Preparation
        - Link to the document: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
2) Create a list of key words that represents the identified key events
        - Any events not defined will be classified as "OTHERS"
3) Not all events occurs singularly, there are events that can occur concurrently, hence affecting the result
        - This is with the assumption that the data is properly documented based on the Storm Data Preparation document
        - As the study is to identify specific events that causes most harm/damages, mix events should be omitted
4) Events are reclassified based on the list of key words
        - Singular events will remain the name of the key event
        - Mix events will have a "MIX EVENT:" place at the front followed by the key event names

```{r Transform Reclassification}

data_event_tidy$KEY = "OTHERS"

# Event reclassification list
events=list()
events["ASTRONOMICAL LOW TIDE"] = list(c("ASTRONOMICAL LOW"))
events["AVALANCHE"]=list(c("AVALA"))
events["BLIZZARD"]=list(c("BLIZZ"))
events["COASTAL FLOOD"]=list(c("COASTAL SURGE", "EROSION CSTL FLOOD", "COASTAL F"))
events["COLD/WIND CHILL"]=list(c("COLD", "LIGHT SNOW"))
events["DEBRIS FLOW"]=list(c("LANDS"))
events["DENSE FOG"]=list(c("DENSE FOG"))
events["DENSE SMOKE"]=list(c("DENSE SMOKE"))
events["DROUGHT"]=list(c("DROUGHT"))
events["DUST DEVIL"]=list(c("DEVIL"))
events["DUST STORM"]=list(c("DUST STORM", "DUST S"))
events["EXCESSIVE HEAT"]=list(c("IVE HEAT", "EXCESSIVE HEAT"))
events["EXTREME COLD/WIND CHILL"]=list(c("EME COLD", "HYPO", "LOW TEM"))
events["FLASH FLOOD"]=list(c("FLASH"))
events["FLOOD"]=list(c("^FLO", "HIGH WA", "FLOODING", "R FLOOD", "N FLOOD", "RISING WATER", "STREAM FLOOD",
                       "STREAM FLD", "FLOOD"))
events["FROST/FREEZE"]=list(c("FROST","FREEZE$", "GLAZE"))
events["FUNNEL CLOUD"]=list(c("FUNNEL CLOUD"))
events["FREEZING FOG"]=list(c("FREEZING FOG"))
events["HAIL"]=list(c("HAIL"))
events["HEAT"]=list(c("HEAT", "Y WARM", "HYPER"))
events["HEAVY RAIN"]=list(c("SIVE RAIN", "HEAVY RAIN", "TORRENTIAL RAIN", "HEAVY PRECIP"))
events["HEAVY SNOW"]=list(c("HEAVY SNOW", "SIVE SNOW", "ING SNOW", "RECORD SNOW", "SNOW SQUALL"))
events["HIGH SURF"]=list(c("HIGH SURF", "HIGH SW", "HIGH WA", "HIGH SU", "HEAVY SU", "ROUGH SEAS", "HAZARDOUS SURF", "HIGH SE"))
events["HIGH WIND"]=list(c("HIGH WIND", "DRY MICRO", "E WIND", "Y WIND"))
events["HURRICANE/TYPHOON"]=list(c("HURRICANE","TYPHOON"))
events["ICE STORM"]=list(c("ICE STORM"))
events["LAKE-EFFECT SNOW"]=list(c("LAKE-EFFECT", "LAKE SNOW", "LAKE EFFECT SNOW"))
events["LAKESHORE FLOOD"]=list(c("LAKESHORE FLOOD", "LAKE FLOOD"))
events["LIGHTNING"]=list(c("LIGHTNING", "LIGHTING", "LIGTNTING"))
events["MARINE HAIL"]=list(c("MARINE HAIL"))
events["MARINE HIGH WIND"]=list(c("MARINE HIGH WIND"))
events["MARINE STRONG WIND"]=list(c("MARINE STRONG WIND"))
events["MARINE THUNDERSTORM WIND"]=list(c("MARINE THUNDERSTORM WIND", "MARINE THUNDERSTORM", "MARINE TSTM WIND"))
events["RIP CURRENT"]=list(c("RIP CURRENT"))
events["SEICHE"]=list(c("SEICHE"))
events["SLEET"]=list(c("SLEET", "MIX", "SPRAY", "ZING RAIN", "ZING DRIZZLE", "RAIN/SNOW")) # Mixed Precipitation
events["STORM SURGE/TIDE"]=list(c("STORM SURGE", "STORM TIDE", "COASTAL STORM", "COASTALSTORM"))
events["STRONG WIND"]=list(c("STRONG WIND", "G WIND", "WIND DAMAGE"))
events["THUNDERSTORM WIND"]=list(c("THUNDERSTORM","THUDERSTORM", "TSTM WIND", "T[A-Z]M WIND", "TSTMW"))
events["TORNADO"]=list(c("TORNADO"))
events["TROPICAL DEPRESSION"]=list(c("TROPICAL DEPRESSION"))
events["TROPICAL STORM"]=list(c("TROPICAL STORM"))
events["TSUNAMI"]=list(c("TSUNAMI"))
events["VOLCANIC ASH"]=list(c("VOLCANANIC ASH"))
events["WATERSPROUT"]=list(c("WATERSPOUT"))
events["WILDFIRE"]=list(c("WILDFIRE", "WILD F", "WILD/FOREST F"))
events["WINTER STORM"]=list(c("WINTER STORM"))
events["WINTER WEATHER"]=list(c("WINTER WEATHER"))


for (i in 1:length(events)){
        flag = 0
        for (j in 1:length(events[[i]])){ 
                data_event_tidy = mutate( data_event_tidy, 
                                          KEY = ifelse( 
                                                  grepl(events[[i]][j], EVTYPE) == TRUE, 
                                                  ifelse(
                                                          KEY == "OTHERS", 
                                                          names(events[i]),
                                                          ifelse(
                                                                  grepl("^MIX EVENTS: ", KEY) == TRUE,
                                                                  ifelse(grepl(names(events[i]), KEY) == TRUE,
                                                                         KEY, paste(KEY, ";", names(events[i]))),
                                                                  ifelse(grepl(names(events[i]), KEY) == TRUE,
                                                                         KEY, paste("MIX EVENTS: ", KEY, ";", 
                                                                                    names(events[i]))))),
                                            KEY))
        }
}

```   

####e. Summarizing the data based on the Key Events
1) Data are summarized based on the reclassified key events

```{r Summary}
data_event_tidy = group_by(data_event_tidy, KEY)
event_summary = summarise(data_event_tidy, 
                       FATALITIES = sum(FATALITIES), 
                       INJURIES = sum(INJURIES), 
                       DAMAGES = sum(DAMAGES))

## Remove old dataset to free up memory
rm(data_event_tidy, events)
```

####f. To check that the bulk of the contribution still lies in the key events
1) The summarized data is further classified to "KEY EVENTS", "NON-KEY EVENTS" and "MIX EVENTS"
        - This is to ensure that the key events are still the main contributor to the fatalities and damages
2) Based on the table, 
        - Mix Event contributes about 
                a. c(event_test[2,5],"%") to the total fatalities 
                b. c(event_test[2,6], "%") to the total damages
        - Non Key Events contributes about
                a. c(event_test[3,5],"%") to the total fatalities 
                b. c(event_test[3,6], "%") to the total damages
3) From the above, it is safe to conclude that Key events are still the main contributor and we can omit non-key events and mix events

```{r Test}
event_summary$identify = ifelse(grepl("MIX", event_summary$KEY) == TRUE, "MIX EVENTS", 
                                ifelse(grepl("OTHERS", event_summary$KEY) == TRUE, "NON-KEY EVENTS", "KEY EVENTS"))
event_test = group_by(event_summary, identify)
event_test = summarise(event_test, 
                       FATALITIES = sum(FATALITIES), 
                       INJURIES = sum(INJURIES), 
                       DAMAGES = sum(DAMAGES))

event_summary = as.data.frame(event_summary[,1:4])   

event_test$PERCENT_FATALITIES = round(prop.table(event_test$FATALITIES)*100,2)
event_test$PERCENT_DAMAGES = round(prop.table(event_test$DAMAGES)*100,2)

event_test

```

####g. To finalize the data for analysis
1) Omit data from mix events and non-key events
2) Summarize data for fatalities and damages

```{r Finalize Data}
## Remove old dataset to free up memory
rm(event_test)

## Highlighting on Single Events 
event_summary$FLAG = ifelse(grepl("MIX", event_summary$KEY) == TRUE | grepl("OTHERS",event_summary$KEY) == TRUE, "1", "0")

## Remove Mix Events
event_summary = filter(event_summary, FLAG == 0)
event_summary = as.data.frame(event_summary[,1:4])        

fatalities_summary = as.data.frame(event_summary[,c(1,2)])
fatalities_summary = arrange(fatalities_summary, desc(FATALITIES))

damages_summary = as.data.frame(event_summary[,c(1,4)])
damages_summary = arrange(damages_summary, desc(DAMAGES))


```


### 2) Analyzing the Data
#### a) Creating the fields for Pareto Plot
```{r Pareto Plot Preparation}
fatalities_summary$CUMM = cumsum(fatalities_summary$FATALITIES)
fatalities_summary$PERCENT_FATALITIES = round(prop.table(fatalities_summary$FATALITIES)*100,2)
fatalities_summary$CUMM_PERCENT_FATAL = cumsum(fatalities_summary$PERCENT_FATALITIES)
fatalities_summary$KEY <- factor(fatalities_summary$KEY, levels=fatalities_summary$KEY)

damages_summary$CUMM = cumsum(damages_summary$DAMAGES)
damages_summary$PERCENT_DAMAGES = round(prop.table(damages_summary$DAMAGES)*100,2)
damages_summary$CUMM_PERCENT_DMG = cumsum(damages_summary$PERCENT_DAMAGES)
damages_summary$KEY <- factor(damages_summary$KEY, levels=damages_summary$KEY)

```


#### b) Plot out the Fatalities Pareto Chart
1) Based on the pareto, there are top 6 events that contributes to ~80% of the total fatalities
2) Details of these 6 events are identified in the table below.

```{r Fatalities_Pareto, fig.path='figure/'}

## Plot out Pareto for Fatalities

g1 = ggplot(fatalities_summary, aes(x = KEY)) 
g2 = geom_bar(aes(y = FATALITIES), fill="red", stat="identity") 
g3 = geom_point(aes(y = CUMM)) 
g4 = geom_path(aes(y = CUMM, group=1))
g5 = theme(axis.text.x = element_text(angle = 90, hjust = 1))
g6 = labs(x = "EVENTS")
g7 = ggtitle('Pareto Plot on Fatalities of Key Events')

g1 + g2 + g3 + g4 + g5 + g6 + g7

top_fatalities = filter(fatalities_summary, CUMM_PERCENT_FATAL <= 80)

top_fatalities

```

#### c) Plot out the Damages Pareto Chart
1) Based on the pareto, there are top 4 events that contributes to ~80% of the total damages
2) Details of these 4 events are identified in the table below.

```{r Damages_Pareto, fig.path='figure/'}

## Plot out Pareto for Fatalities

g1 = ggplot(damages_summary, aes(x = KEY)) 
g2 = geom_bar(aes(y = DAMAGES), fill="red", stat="identity") 
g3 = geom_point(aes(y = CUMM)) 
g4 = geom_path(aes(y = CUMM, group=1))
g5 = theme(axis.text.x = element_text(angle = 90, hjust = 1))
g6 = labs(x = "EVENTS")
g7 = ggtitle('Pareto Plot on Damages of Key Events')

g1 + g2 + g3 + g4 + g5 + g6 + g7

top_damages = filter(damages_summary, CUMM_PERCENT_DMG <= 80)

top_damages

```



### End of Report.
